---
title: "Macrophages LPS estimulation genelists. 5x10-2 pval"
output:
  html_document:
  df_print: paged
---

# Info about the files


This analysis uses the Supplementary data from Panousis et al (2023), that can be found in https://www.biorxiv.org/content/10.1101/2023.05.29.542425v1.supplementary-material.
In Suppl table 1 there is some technical information. 
Suppl table 2 has the data of the stimulation at 6h and Suppl. table 3 at 24h.
For this analysis I first created csv files for each stimuli and timepoint that are used as input.


```{r}
library("EnsDb.Hsapiens.v86")
library(UpSetR)
library(gplots)
library(VennDiagram)
```

# Predefined functions

## Display Venn Diagram
```{r}
# Helper function to display Venn diagram
display_venn <- function(x, ...){
  grid.newpage()
  venn_object <- venn.diagram(x, disable.logging = FALSE,
                              filename = NULL, ...)
  grid.draw(venn_object)
}
```


## GeneID converter
```{r}
geneID_converter <- function(goi)
  {
  #goi <- AGGlps_Mac7_Mlps_LR$gene
  ens_df <- ensembldb::select(EnsDb.Hsapiens.v86, keys= goi, 
                              keytype = "SYMBOL", 
                              columns = c("SYMBOL","GENEID"))
  ens_df <- ens_df$GENEID[grepl(pattern = "^ENS", x = ens_df$GENEID)]
  return(ens_df)
  }
```

## Read and filter input
```{r}
read_DE_list <- function(infile)
{ 
  df <- read.csv(file = infile,sep = ";")
  df <- df[df$log2FoldChange>1, ]
  df <- df[df$padj < 0.05, ]  
  df <- df[order(df$log2FoldChange, decreasing = T),]
  df <- df[complete.cases(df), ]
  #ENS_ID <- geneID_converter(df$hgnc)
  #return(ENS_ID)
  return(df)
}
```


## Write output lists 
```{r}
out_folder <- "./Output_lists/pval5x10-2_sorted/"
get_DE_set_lists <- function(after6h, after24h, stimuli_name)
{
  early <- setdiff(after6h$hgnc, after24h$hgnc)
  inter <- intersect(after6h$hgnc, after24h$hgnc)
  late  <- setdiff(after24h$hgnc, after6h$hgnc)

  Early_genes <<- rbind(Early_genes, after6h[after6h$hgnc%in%early, ])
  Inter_genes <<- rbind(Inter_genes, after6h[after6h$hgnc%in%inter, ],
                        after24h[after24h$hgnc%in%inter, ])
  Late_genes  <<- rbind(Late_genes, after24h[after24h$hgnc%in%late, ])
  
#  write(x = early, file = paste0(out_folder,stimuli_name,"sorted_early.txt"))
#  write(x = inter, file = paste0(out_folder,stimuli_name,"sorted_inter.txt"))
#  write(x = late,  file = paste0(out_folder,stimuli_name,"sorted_late.txt"))
}
```

# Lists for Early, Inter and Late

```{r}
df_colnames <- c("hgnc","Gene","baseMean","log2FoldChange",
                 "lfcSE","stat","pvalue","padj")
Early_genes <- as.data.frame(matrix(data = NA, nrow = 0, ncol = 8))
colnames(Early_genes) <- df_colnames
Inter_genes <- Early_genes
Late_genes  <- Early_genes
```


# Stimuli 

## IFNB: Interferon beta, anti-inflammatory macrophages

```{r, fig.width=4, fig.height=4}
IFNB_6hrs <- read_DE_list("Stimuli_lists/Ctrl_vs_IFNB (6 hours).csv")
IFNB_24hrs <- read_DE_list("Stimuli_lists/Ctrl_vs_IFNB (24 hours).csv")
display_venn(list(IFNB_6hrs = IFNB_6hrs$hgnc, IFNB_24hrs= IFNB_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = IFNB_6hrs, after24h = IFNB_24hrs, stimuli_name = "IFNB")
```

##  IFNG: Interferon gamma, pro-inflammatory macrophages

```{r, fig.width=4, fig.height=4}
IFNG_6hrs <- read_DE_list("Stimuli_lists/Ctrl_vs_IFNG (6 hours).csv")
IFNG_24hrs <- read_DE_list("Stimuli_lists/Ctrl_vs_IFNG (24 hours).csv")

display_venn(list(IFNG_6hrs = IFNG_6hrs$hgnc, IFNG_24hrs= IFNG_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = IFNG_6hrs, after24h = IFNG_24hrs, stimuli_name = "IFNG")
```
## IL4: interleukin 4, anti-inflammatory macrophages

```{r, fig.width=4, fig.height=4}
IL4_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_IL4 (6 hours).csv")
IL4_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_IL4 (24 hours).csv")

display_venn(list( IL4_6hrs=IL4_6hrs$hgnc ,IL4_24hrs = IL4_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = IL4_6hrs, after24h = IL4_24hrs, stimuli_name = "IL4")
```
## R484: resquimoid, TLR 7/8 agonist

```{r, fig.width=4, fig.height=4}
R484_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_R848 (6 hours).csv")
R484_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_R848 (24 hours).csv")

display_venn(list(R484_6hrs =R484_6hrs$hgnc ,R484_24hrs =R484_24hrs$hgnc ), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h =R484_6hrs , after24h = R484_24hrs, stimuli_name = "R484")
```

## PIC: poly I:C, similar to dsRNA, mimics exposure to viruses


```{r, fig.width=4, fig.height=4}
PIC_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_PIC (6 hours).csv")
PIC_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_PIC (24 hours).csv")

display_venn(list(PIC_6hrs =PIC_6hrs$hgnc ,PIC_24hrs = PIC_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = PIC_6hrs, after24h = PIC_24hrs, stimuli_name = "PIC")
```


## sLPS: Salmonella typhimurium lipopolysaccharide, pro-inflammatory macrophages

```{r, fig.width=4, fig.height=4}
sLPS_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_sLPS (6 hours).csv")
sLPS_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_sLPS (24 hours).csv")

display_venn(list(sLPS_6hrs =sLPS_6hrs$hgnc , sLPS_24hrs= sLPS_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = sLPS_6hrs, after24h = sLPS_24hrs, stimuli_name = "sLPS")
```

## P3C: Pam3CSK4, Pam3CysSerLys4 TLR1/2 agonist

```{r, fig.width=4, fig.height=4}
P3C_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_P3C (6 hours).csv")
P3C_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_P3C (24 hours).csv")

display_venn(list(P3C_6hrs =P3C_6hrs$hgnc , P3C_24hrs=P3C_24hrs$hgnc ), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = P3C_6hrs, after24h = P3C_24hrs, stimuli_name = "P3C")
```


## CIL: CD40 ligand and IFNγ + sLPS

```{r, fig.width=4, fig.height=4}
CIL_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_CIL (6 hours).csv")
CIL_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_CIL (24 hours).csv")

display_venn(list( CIL_6hrs= CIL_6hrs$hgnc, CIL_24hrs= CIL_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = CIL_6hrs, after24h = CIL_24hrs, stimuli_name = "CIL")
```


## LIL10: interleukin 10 + LPS, anti-inflamatory effect of IL10 in LPS stimulated cells

```{r, fig.width=4, fig.height=4}
LIL10_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_LIL10 (6 hours).csv")
LIL10_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_LIL10 (24 hours).csv")

display_venn(list(LIL10_6hrs =LIL10_6hrs$hgnc ,LIL10_24hrs =LIL10_24hrs$hgnc ), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = LIL10_6hrs, after24h = LIL10_24hrs, stimuli_name = "LIL10")
```
## MBP: Myelin basic protein, damage induced by multiple sclerosis

```{r, fig.width=4, fig.height=4}
MBP_6hrs = read_DE_list("Stimuli_lists/Ctrl_vs_MBP (6 hours).csv")
MBP_24hrs = read_DE_list("Stimuli_lists/Ctrl_vs_MPB (24 hours).csv")

display_venn(list(MBP_6hrs =MBP_6hrs$hgnc ,MBP_24hrs = MBP_24hrs$hgnc), 
             fill = c("tomato", "seagreen" ))
get_DE_set_lists(after6h = MBP_6hrs, after24h = MBP_24hrs, stimuli_name = "MBP")
```


# Final lists 

```{r, eval=TRUE}
Early_genes <- Early_genes[order(Early_genes$log2FoldChange, decreasing = TRUE), ]
Inter_genes <- Inter_genes[order(Inter_genes$log2FoldChange, decreasing = TRUE), ]
Late_genes  <- Late_genes[order(Late_genes$log2FoldChange, decreasing = TRUE), ]

Early_genes_unique <- Early_genes[!duplicated(Early_genes$hgnc), ]
Inter_genes_unique <- Inter_genes[!duplicated(Inter_genes$hgnc), ]
Late_genes_unique  <- Late_genes[!duplicated(Late_genes$hgnc), ]

Early_genes_ENS <- geneID_converter(Early_genes_unique$hgnc)
Inter_genes_ENS <- geneID_converter(Inter_genes_unique$hgnc)
Late_genes_ENS  <- geneID_converter(Late_genes_unique$hgnc)
```


```{r, eval=FALSE}
write(x = Early_genes_ENS, file = paste0(out_folder,"Early_genes_sorted.txt"))
write(x = Inter_genes_ENS, file = paste0(out_folder,"Inter_genes_sorted.txt"))
write(x = Late_genes_ENS,  file = paste0(out_folder,"Late_genes_sorted.txt"))

```


```{r}

display_venn(list(early = Early_genes_ENS ,inter = Inter_genes_ENS, late= Late_genes_ENS), 
             fill = c("tomato", "seagreen", "blue" ))
```

```{r}
sessionInfo()

```

