---
title:  "PS enrichment Developmental Atlas. HyperGeometric test"
author: "Irepan Salvador-Martinez"
date: "14/07/2023"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
---

Here I take the top 500 DE genes for each celltype (same we used for estimating
alpha) and test if these genes are enriched with a specific Phylostrata.

For this I used the dataset from Litman and Stein where they asign a PS value
to all human genes using the median/mean PS value of different datasets



```{r}
library(gplots)
library(RColorBrewer)
library(scico)
library(reshape2)
library(ggplot2)

```


```{r}

process_data <- function(data, lineage) {
  df <- melt(data)
  colnames(df) <- c("celltype", "Phylostrata", "pval")
  df$celltype <- stringr::str_remove_all(df$celltype, "_FULL_ENSEMBL.txt")
  df$pval[df$pval < 1e-9 ] = 1e-9  
  df$lineage <- lineage
  return(df)
}


plot_HM <- function(data, title)
{
  ggplot(data, aes(x = Phylostrata, y = celltype, fill = pval)) +
  geom_tile() +
  colorspace::scale_fill_continuous_divergingx(
    limits = c(1e-9, 1),
    breaks=c(1e-9, 1e-6, 1e-3, 1),
    trans = "log10",
    name = "p-value",
    mid = log10(0.001),
    palette = "ArmyRose"
  ) +
  theme_classic() + ggtitle(title) +
  scale_x_continuous("Phylostrata", c(1:19) ) +
  scale_y_discrete("celltype")
}



```



```{r}
metadata <- read.csv(file = "Human_Gene_Ages.csv")

Gene_ages <- metadata[,c(1,7)]

Gene_ages$Gene_age <- as.factor(Gene_ages$Gene_age)
plot(table(Gene_ages$Gene_age), type = "line")

```
# ALL PHYLOSTRATA


## Progenitors

```{r}
in_dir <- "../Developmental/ENS_FULL_genelists_wilcox/HSC_progenitors/"
files <- list.files(path = "../Developmental/ENS_FULL_genelists_wilcox/HSC_progenitors/", pattern = "*.txt")
N <- length(Gene_ages$Gene_age)
N_PS <- table(Gene_ages$Gene_age)

Hyper_test_Prog <- matrix(data = NA, nrow = length(files), ncol = length(N_PS),
                     dimnames = list(files,1:length(N_PS)) )
Prog_mean_Age <-  data.frame(celltype = files, Mean_Age = NA )

for (i in files) {
  print(i)
  goi <- read.csv(file = paste0(in_dir, i))
  goi <- goi[, 1]
  if (length(goi) > 500) {goi <- goi[1:500] } # top 500

  ages <- Gene_ages[which(Gene_ages$Gene_id %in% goi), ]
  #if (length(ages$Gene_id) > 500) {ages <- ages[1:500,] } # top 500
  n <- length(ages$Gene_age)
  n_ps <- table(ages$Gene_age)
  Prog_mean_Age$Mean_Age[Prog_mean_Age$celltype == i] <- mean(as.numeric(ages$Gene_age))
  
  for (j in 1:length(N_PS)) {
    #j <- 1
    # q number of genes of a PS in a sample
    # m the number of genes of a PS in the pop
    # n the number of genes in the population that are not of a PS
    # k the number genes in the sample (all PS)
    pval <- phyper(q = n_ps[j]-1, m = N_PS[j], n = N - N_PS[j], k = n,
                   lower.tail = FALSE)
    Hyper_test_Prog[i,j] <- pval  
  }
}
```

```{r}
Progen <- process_data(Hyper_test_Prog, "Progenitors")
Progen$celltype <- paste0(Progen$celltype, "_progen")
plot_HM(Progen, "Progenitors")
```


## Myeloid

```{r}
in_dir <- "../Developmental/ENS_FULL_genelists_wilcox/Myeloid_ALL/"
files <- list.files(path = "../Developmental/ENS_FULL_genelists_wilcox/Myeloid_ALL/", pattern = "*.txt")
N <- length(Gene_ages$Gene_age)
N_PS <- table(Gene_ages$Gene_age)

Hyper_test_Myel <- matrix(data = NA, nrow = length(files), ncol = length(N_PS),
                     dimnames = list(files,1:length(N_PS)) )
Myel_mean_Age <-  data.frame(celltype = files, Mean_Age = NA )
for (i in files) {
  print(i)
  goi <- read.csv(file = paste0(in_dir, i))
  goi <- goi[, 1]
  if (length(goi) > 500) {goi <- goi[1:500] } # top 500

  ages <- Gene_ages[which(Gene_ages$Gene_id %in% goi), ]
  #if (length(ages$Gene_id) > 500) {ages <- ages[1:500,] } # top 500
  n <- length(ages$Gene_age)
  n_ps <- table(ages$Gene_age)
  Myel_mean_Age$Mean_Age[Myel_mean_Age$celltype == i] <- mean(as.numeric(ages$Gene_age))
  
  for (j in 1:length(N_PS)) {
    #j <- 1
    # q number of genes of a PS in a sample
    # m the number of genes of a PS in the pop
    # n the number of genes in the population that are not of a PS
    # k the number genes in the sample (all PS)
    pval <- phyper(q = n_ps[j]-1, m = N_PS[j], n = N - N_PS[j], k = n,
                   lower.tail = FALSE)
    Hyper_test_Myel[i,j] <- pval  
  }
}
```

```{r}
Myeloid <- process_data(Hyper_test_Myel, "Myeloid")
plot_HM(Myeloid, "Myeloid")
```


## Lymphoid

```{r}
in_dir <- "../Developmental/ENS_FULL_genelists_wilcox/Lymphoid_ALL/"
files <- list.files(path = "../Developmental/ENS_FULL_genelists_wilcox/Lymphoid_ALL/", pattern = "*.txt")
N <- length(Gene_ages$Gene_age)
N_PS <- table(Gene_ages$Gene_age)

Hyper_test_Lymp <- matrix(data = NA, nrow = length(files), ncol = length(N_PS),
                     dimnames = list(files,1:length(N_PS)) )
Lymp_mean_Age <-  data.frame(celltype = files, Mean_Age = NA )
for (i in files) {
  print(i)
  goi <- read.csv(file = paste0(in_dir, i))
  goi <- goi[, 1]
  if (length(goi) > 500) {goi <- goi[1:500] } # top 500

  ages <- Gene_ages[which(Gene_ages$Gene_id %in% goi), ]
  #if (length(ages$Gene_id) > 500) {ages <- ages[1:500,] } # top 500
  n <- length(ages$Gene_age)
  n_ps <- table(ages$Gene_age)
  Lymp_mean_Age$Mean_Age[Lymp_mean_Age$celltype == i] <- mean(as.numeric(ages$Gene_age))
  
  for (j in 1:length(N_PS)) {
    #j <- 1
    # q number of genes of a PS in a sample
    # m the number of genes of a PS in the pop
    # n the number of genes in the population that are not of a PS
    # k the number genes in the sample (all PS)
    pval <- phyper(q = n_ps[j]-1, m = N_PS[j], n = N - N_PS[j], k = n,
                   lower.tail = FALSE)
    Hyper_test_Lymp[i,j] <- pval  
  }
}
```
```{r}
Lymphoid <- process_data(Hyper_test_Lymp, "Lymphoid")
plot_HM(Lymphoid, "Lymphoid")
```

## MegaK_Ery

```{r}
in_dir <- "../Developmental/ENS_FULL_genelists_wilcox/MegaK_Ery/"
files <- list.files(path = "../Developmental/ENS_FULL_genelists_wilcox/MegaK_Ery/", pattern = "*.txt")
N <- length(Gene_ages$Gene_age)
N_PS <- table(Gene_ages$Gene_age)

Hyper_test_MK_Ery <- matrix(data = NA, nrow = length(files), ncol = length(N_PS),
                     dimnames = list(files,1:length(N_PS)) )
MK_Ery_mean_Age <-  data.frame(celltype = files, Mean_Age = NA )

for (i in files) {
  print(i)
  goi <- read.csv(file = paste0(in_dir, i))
  goi <- goi[, 1]
  if (length(goi) > 500) {goi <- goi[1:500] } # top 500

  ages <- Gene_ages[which(Gene_ages$Gene_id %in% goi), ]
  #if (length(ages$Gene_id) > 500) {ages <- ages[1:500,] } # top 500
  n <- length(ages$Gene_age)
  n_ps <- table(ages$Gene_age)
  MK_Ery_mean_Age$Mean_Age[MK_Ery_mean_Age$celltype == i] <- mean(as.numeric(ages$Gene_age))
  
  for (j in 1:length(N_PS)) {
    #j <- 1
    # q number of genes of a PS in a sample
    # m the number of genes of a PS in the pop
    # n the number of genes in the population that are not of a PS
    # k the number genes in the sample (all PS)
    pval <- phyper(q = n_ps[j]-1, m = N_PS[j], n = N - N_PS[j], k = n,
                   lower.tail = FALSE)
    Hyper_test_MK_Ery[i,j] <- pval  
  }
}
```

```{r}
MK_Ery <- process_data(Hyper_test_MK_Ery, "MegaK/Erythrocytes")
plot_HM(MK_Ery, "MegaK/Erythrocytes")
```



```{r}
All_celltypes <- rbind(Progen, Myeloid, Lymphoid, MK_Ery)
All_celltypes$celltype <- factor(All_celltypes$celltype, levels = unique(All_celltypes$celltype))
plot_HM(All_celltypes, "Developmental Atlas")
```

```{r}
for (i in unique(All_celltypes$Phylostrata))
{
  print(i)
  x <- All_celltypes[All_celltypes$Phylostrata == i,]
  y <- sum(x$pval <= 0.001)/length(x$pval)
  print(y)  
}
```


## export mean ages tables

```{r, eval=FALSE}
write.csv(MK_Ery_mean_Age, file = "./Mean_Ages/MK_Ery_mean_Age.csv")
write.csv(Lymp_mean_Age, file = "./Mean_Ages/Lymp_mean_Age.csv")
write.csv(Myel_mean_Age, file = "./Mean_Ages/Myel_mean_Age.csv")
write.csv(Prog_mean_Age, file = "./Mean_Ages/Prog_mean_Age.csv")
```


```{r}
sessionInfo()
```

